Highly Available Transactions: Virtues and Limitations

### 2013-03-31 writeup/evaluation.tex
checkpoint

\section{Application and System Analysis}
\label{sec:evaluation}

@@

\vspace{.5em}\noindent\textbf{Serializablity} To establish a serial
order on transactions, algorithms for achieving serializability of
general-purpose read-write transactions in a distributed
setting~\cite{bernstein-book} may require at least one round trip time
(RTT) before committing. As an example, traditional two-phase locking
for a transaction of length $T$ requires $T$ \texttt{lock} operations
and at least one \texttt{unlock} operation.  In a distributed
environment, each of these operations requires coordination, either
with other database replicas or with a lock service. If the
coordination mechanism for locking is unavailable, transactions cannot
commit. Similarly, optimistic concurrency control requires
coordinating via a validation step, while deterministic transaction
scheduling~\cite{deterministic-scheduling} requires a
scheduler. Serializability under multi-version concurrency control
also requires checking for update conflicts.

@@

\vspace{.5em}\noindent\textbf{Non-serializablity} Many existing
distributed implementations of weak isolation are not highly
available. Lock-based proposals such as those used to provide weak
isolation in Gray's original proposal~\cite{gray-isolation} do not
degrade gracefully in the presence of partial failure. (Note, however,
that lock-based protocols \textit{do} offer the benefit of recency
guarantees.) While multi-versioned storage systems allow for a variety
of transactional guarantees, we have not seen proposals for truly weak
isolation (e.g., non-``tentative update'' schemes) in this
context. Chan and Gray's read-only transactions provide read-only
transactions with item-cut isolation, causal consistency, and
transactional atomicity (session PL-2L) but are unavailable in the
presence of coordinator failure~\cite{readonly}, similar to read-only
and write-only transactions more recently proposed by
Eiger~\cite{eiger}.  Bolt-on causal consistency~\cite{bolton},
Brantner's S3 database~\cite{kraska-s3}, and
Bayou~\cite{sessionguarantees} can all provide variants of session
PL-2L with high availability, but none provide this HAT functionality
without substantial modification; Swift~\cite{swift} is closest to
providing full HAT operation in a sticky context. Similarly, the
MDCC~\cite{mdcc} protocol offers Read Committed isolation with Lost
Update avoidance but is similarly unavailable due to its reliance on
preventing write conflicts.

>>>>

\section{HAT Implications}
\label{sec:evaluation}

@@

\vspace{.5em}\noindent\textbf{Serializability} To establish a serial
order on transactions, algorithms for achieving serializability of
general-purpose read-write transactions in a distributed
setting~\cite{bernstein-book} may require at least one round trip time
(RTT) before committing. As an example, traditional two-phase locking
for a transaction of length $T$ requires $T$ \texttt{lock} operations
and at least one \texttt{unlock} operation.  In a distributed
environment, each of these operations requires coordination, either
with other database replicas or with a lock service. If the
coordination mechanism for locking is unavailable, transactions cannot
commit. Similarly, optimistic concurrency control requires
coordinating via a validation step, while deterministic transaction
scheduling~\cite{deterministic-scheduling} requires a
scheduler. Serializability under multi-version concurrency control
also requires checking for update conflicts.

@@

\vspace{.5em}\noindent\textbf{Non-serializability} Many existing
distributed implementations of weak isolation are not highly
available. Lock-based proposals such as those used to provide weak
isolation in Gray's original proposal~\cite{gray-isolation} do not
degrade gracefully in the presence of partial failure. (Note, however,
that lock-based protocols \textit{do} offer the benefit of recency
guarantees.) While multi-versioned storage systems allow for a variety
of transactional guarantees, we have not seen proposals for truly weak
isolation (e.g., non-``tentative update'' schemes) in this
context. Chan and Gray's read-only transactions provide read-only
transactions with item-cut isolation, causal consistency, and
transactional atomicity (session PL-2L) but are unavailable in the
presence of coordinator failure~\cite{readonly}, similar to read-only
and write-only transactions more recently proposed by
Eiger~\cite{eiger}.  Bolt-on causal consistency~\cite{bolton},
Brantner's S3 database~\cite{kraska-s3}, and
Bayou~\cite{sessionguarantees} can all provide variants of session
PL-2L with high availability, but none provide this HAT functionality
without substantial modification; Swift~\cite{swift} is closest to
providing full HAT operation in a sticky context. Similarly, the
MDCC~\cite{mdcc} protocol offers Read Committed isolation with Lost
Update avoidance but is similarly unavailable due to its reliance on
preventing write conflicts.

### 2013-03-31 writeup/hat-analysis.tex
checkpoint

When possible, we draw on existing properties and definitions from the
database and distributed systems literature, providing a brief,
informal explanation and example for each guarantee. In particular,
for our ACID guarantees, we draw largely on Atul Adya's
dissertation~\cite{adya}, Berenson et al.'s 1995 critique of the ANSI
SQL Specification~\cite{ansicritique}, and the ANSI SQL
specification~\cite{ansi-sql}. We provide a set of formal definitions
and semantics in our extended Technical Report~\cite{hat-tr}.

>>>>

As Brewer states, ``systems and database communities are separate but
overlapping (with distinct vocabulary)''~\cite{brewer-slides}. With
this challenge in mind, we draw on existing properties and definitions
from the database and distributed systems literature, providing a
brief, informal explanation and example for each guarantee. In
particular, for our ACID guarantees, we draw largely on Atul Adya's
dissertation~\cite{adya}, Berenson et al.'s 1995 critique of the ANSI
SQL Specification~\cite{ansicritique}, and the ANSI SQL
specification~\cite{ansi-sql}. We provide a set of formal definitions
and semantics in our extended Technical Report~\cite{hat-tr}.

### 2013-03-31 writeup/intro.tex
checkpoint

The prior decade saw enormous churn in the design of widely-deployed
database backends, from the de facto use of transactional
RDBMSs~\cite{bernstein-concurrency, bernstein-book, gray-isolation,
  gray-virtues} to the widespread adoption of loosely-consistent
distributed key-value stores~\cite{bigtable, pnuts, dynamo}, with
increasing interest in returning to transactional
designs~\cite{spanner, walter, foundation-article, krikellas-bargain,
  eiger}. Instrumental in these trends was the 2000 introduction of
Brewer's CAP Theorem, which stated that a highly available system
cannot provide strong consistency guarantees in the presence of
network partitions~\cite{brewer-slides}. For distributed systems
designers with requirements for ``always on,'' highly available and
low latency operation, CAP became an easily understandable indication
of the competing goals of semantic strength and availability. However,
as formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability, the ability to read the most recent write to a data
item that is replicated across servers. Yet this result---regarding a
fairly narrow form of distributed consistency---is often conflated
with ACID database properties~\cite{brewer-slides, hn,
  foundation-article}, leading to substantial confusion about the
relationships between replica consistency, transactional isolation, and high
availability.

Serializability, the gold standard of traditional ACID transactions,
is indeed not achievable given a requirement for high availability in
the presence of partitions~\cite{davidson-survey}. However, database
systems have a long tradition of providing weaker isolation and
consistency guarantees~\cite{adya, ansicritique, gray-isolation,
  gray-virtues}. Due to their concurrency and overall performance
benefits, weak isolation models are widely employed, overwhelmingly as
the default setting in today's traditional ``ACID'' and up-and-coming
``NewSQL'' databases and often as the maximum guarantee offered
(Section~\ref{sec:modernacid}). While weak isolation levels do not
provide serializability for general-purpose transactions, they are
apparently strong enough to deliver acceptable behavior to many
application programmers and are substantially stronger than the
semantics of current highly available systems. This raises a natural
question: which of these alaternative semantics can be provided while
remaining highly available?

@@

The contributions of this paper fill the gap between, on one hand, the
plethora of previously proposed isolation and consistency models, and,
on the other hand, the goal of high availability. We introduce a
hierarchy of consistency models and, in doing so, unify previous work
on ACID properties~\cite{adya}, session
guarantees~\cite{sessionguarantees}, and traditional distributed
``register'' semantics~\cite{herlihy-art}. We rigorously taxonomize
the availability characteristics of these consistency models: we
determine which can be offered as Highly Available Transactions
(HATs). In doing so, we demonstrate that many \textit{implementations}
that provide these semantics are not highly available but that their
unavailability is not \textit{inherently} required.

Our investigation shows, somewhat surprisingly, that besides
serializability, only fairly strong models like Snapshot Isolation
(SI) and Repeatable Reads (RR) are not highly available, stemming from
their requirement to detect conflicts between concurrent updates (as
needed for preventing Lost Updates or Write Skew phenomena). In
contrast, Read Committed isolation, Transactional Atomicity, and many
of the weaker data consistency models from database and distributed
systems can be implemented in a highly available manner. For several
of these guarantees, such as Transactional Causal Predicate Cut
Isolation, we consider a modified form of high availiabity in which
clients ``stick to'' (i.e. have affinity with) at least one server (a
property which is often implicit in the distributed systems
literature~\cite{herlihy-art, cops, eiger, cac} but which requires
\textit{explicit} consideration in a client-server replicated database
context). This sticky availability is easily implementatable and
widely employed~\cite{cops, eiger, vogels-defs} but is strictly less
available than traditional definitions of high availability.

>>>>

The prior decade saw a large shift in the design of large-scale
database backends, from the use of transactional
RDBMSs~\cite{bernstein-concurrency, bernstein-book, gray-isolation,
  gray-virtues} to the widespread adoption of loosely-consistent
distributed key-value stores~\cite{bigtable, pnuts,
  dynamo}. Instrumental in this trend was the 2000 introduction of
Brewer's CAP Theorem, which stated that a highly available system
cannot provide strong consistency guarantees in the presence of
network partitions~\cite{brewer-slides}. However, as formally
proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability~\cite{herlihy-art}, the ability to read the most
recent write to a data item that is replicated across servers. Yet
this result---regarding a fairly narrow form of distributed
consistency---is often conflated with ACID database
properties~\cite{brewer-slides, hn, foundation-article}, leading to
substantial confusion about the relationships between replica
consistency, transactional isolation, and high availability. However,
transactions \textit{are} useful for programmers as evidenced by the
recent resurgence of transactional data stores such as Google's
Spanner~\cite{spanner}. Unfortunately, these systems are not highly
available in the presence of partitions~\cite{walter, mdcc,
  foundation-article, krikellas-bargain, eiger}. (TODO:ADD more)

Indeed, serializable transactions, the gold standard of traditional
ACID databases, is indeed not achievable given a requirement for high
availability in the presence of
partitions~\cite{davidson-survey}. However, database systems have a
long tradition of providing weaker isolation and consistency
guarantees~\cite{adya, ansicritique, gray-isolation,
  gray-virtues} (TODO: kemme). Due to their concurrency and overall performance
benefits, weak isolation models are widely employed, overwhelmingly as
the default setting in today's traditional ``ACID'' and up-and-coming
``NewSQL'' databases and often as the maximum guarantee offered
(Section~\ref{sec:modernacid}). While weak isolation levels do not
provide serializability for general-purpose transactions, they are
apparently strong enough to deliver acceptable behavior to many
application programmers and are substantially stronger than the
semantics of current highly available systems. This raises a natural
question: which of these alaternative semantics can be provided while
remaining highly available?

@@

Our main contributions are as follows. We fill gap between, on one
hand, the plethora of previously proposed isolation and consistency
models, and, on the other hand, the goal of high availability.  We
classify which, among the wide array models, are highly available,
denoting them as {\em Highly Available Transactions (HATs)}. In doing
so, we demonstrate that many \textit{implementations} of HAT semantics
are not highly available but that their unavailability is not
\textit{inherently} required. Our investigation shows, somewhat
surprisingly, that besides serializability, only Snapshot Isolation
(SI) and Repeatable Reads (RR) are unavailable. We also show that many
of the weaker data consistency models from distributed systems are
HAT-compliant.

More precisely, highly available systems cannot detect conflicts
between concurrent updates (as needed for preventing Lost Updates or
Write Skew phenomena). However, Read Committed isolation,
Transactional Atomicity, and many of the weaker data consistency
models from database and distributed systems can be implemented in a
highly available manner. For several of these guarantees, such as
Transactional Causal Predicate Cut Isolation, we consider a modified
form of high availiabity in which clients ``stick to'' (i.e., have
affinity with) at least one server (a property which is often implicit
in the distributed systems literature~\cite{herlihy-art, cops, eiger,
  cac} but which requires \textit{explicit} consideration in a
client-server replicated database context). This sticky availability
is easily implementatable and widely employed~\cite{cops, eiger,
  vogels-defs} but is strictly less available than traditional
definitions of high availability.

### 2013-03-31 writeup/motivation.tex
Couple very minor fixes
checkpoint

\section{Motivation}
\label{sec:motivation}

@@

\begin{table}
\subfloat[Cross-region (OR:~Oregon, VA:~Virginia, TO:~Tokyo, IR:~Ireland, SY:~Sydney, SP:~S\~{a}o Paulo, SI:~Singapore)] {
  \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
& \multicolumn{1}{c}{OR} & \multicolumn{1}{c}{VA} & \multicolumn{1}{c}{TO} & \multicolumn{1}{c}{IR} & \multicolumn{1}{c}{SY} & \multicolumn{1}{c}{SP} & \multicolumn{1}{c|}{SI} \\\hline
CA & \colorbox{min-lat-color}{22.5}   & 84.5   & 143.7   & 169.8   & 179.1   & 185.9   & 186.9  \\
OR &  & 82.9   & 135.1   & 170.6   & 200.6   & 207.8   & 234.4  \\
VA & &  & 202.4   & 107.9   & 265.6   & 163.4   & 253.5  \\
TO & & &  & 278.3   & 144.2   & 301.4   & 90.6  \\
IR & & & &  & 346.2   & 239.8   & 234.1  \\
SY & & & & &  & 333.6   & 243.1  \\
SP & & & & & &  & \colorbox{max-lat-color}{362.8}  \\
\hline
  \end{tabular}
}\vspace{.5em}

@@

\subsection{CAP, Isolation, and Modern Systems}
\label{sec:modernacid}

@@

One of the most popularly celebrated impossibility results in
distributed systems is Brewer's CAP Theorem, which, as formally proven
by Gilbert and Lynch, states that it is impossible to maintain
linearizability , or the ability to read the last completed write,
(``C,'' or ``consistency''---albeit poorly named) and replica
availability, or a guaranteed response, (``A'') in the presence of
network partitions (``P'')~\cite{gilbert-cap}. Brewer's Theorem has
roots in decades of distributed database
research~\cite{davidson-survey} and has had a substantial impact on
recent large-scale distributed systems design: while CAP is narrowly
scoped to the impossibility of building highly available linearizable
systems, it is often interpreted to preclude a larger set of
guarantees, including ACID (transactional atomicity, consistency,
isolation and durability guarantees)~\cite{hn, foundation-article} as
also evidenced by the lack of highly available systems providing them.

@@

It is unknown \textit{which} of these ACID (particularly isolation)
guarantees can be provided with high availability. The existing
algorithms for providing weak isolation are typically designed for a
single-node context and are, to the best of our knowledge, unavailable
due to a reliance on concurrency control mechanisms like locking that
are not resilient to partial failure
(Section~\ref{sec:evaluation}). Morever, we are not aware of any prior
literature that provides guidance as to the relationship between weak
isolation and high availability: prior work has examined the
relationship between serializability and high
availability~\cite{davidson-survey} and weak isolation on a
single-server~\cite{adya, ansicritique} but never weak isolation and
high availability taken together.

>>>>

\section{Partitions and Latency}
\label{sec:motivation}

@@

\begin{table}
\subfloat[Cross-region (CA:~California, OR:~Oregon, VA:~Virginia, TO:~Tokyo, IR:~Ireland, SY:~Sydney, SP:~S\~{a}o Paulo, SI:~Singapore)] {
  \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
& \multicolumn{1}{c}{OR} & \multicolumn{1}{c}{VA} & \multicolumn{1}{c}{TO} & \multicolumn{1}{c}{IR} & \multicolumn{1}{c}{SY} & \multicolumn{1}{c}{SP} & \multicolumn{1}{c|}{SI} \\\hline
CA & \colorbox{min-lat-color}{22.5}   & 84.5   & 143.7   & 169.8   & 179.1   & 185.9   & 186.9  \\
OR &  & 82.9   & 135.1   & 170.6   & 200.6   & 207.8   & 234.4  \\
VA & &  & 202.4   & 107.9   & 265.6   & 163.4   & 253.5  \\
TO & & &  & 278.3   & 144.2   & 301.4   & 90.6  \\
IR & & & &  & 346.2   & 239.8   & 234.1  \\
SY & & & & &  & 333.6   & 243.1  \\
SP & & & & & &  & \colorbox{max-lat-color}{362.8}  \\
\hline
  \end{tabular}
}\vspace{.5em}

@@

\section{CAP and Modern Systems}
\label{sec:modernacid}

@@

One of the most popularly celebrated impossibility results in
distributed systems is Brewer's CAP Theorem~\cite{brewer-slides},
which, as formally proven by Gilbert and Lynch~\cite{gilbert-cap},
states that it is impossible to maintain linearizability , or the
ability to read the last completed write, (``C,'' or
``consistency''---albeit poorly named) and replica availability, or a
guaranteed response, (``A'') in the presence of network partitions
(``P''). Brewer's Theorem has roots in decades of distributed database
research~\cite{davidson-survey} and has had a substantial impact on
recent large-scale distributed systems design~\cite{bigtable, pnuts,
  dynamo}: while CAP is narrowly scoped to the impossibility of
building highly available linearizable systems, it is often
interpreted to preclude a larger set of guarantees, including ACID
(transactional atomicity, consistency, isolation and durability
guarantees). We are unlikely to determine the exact cause of this
confusion, but it appears as early as the first public announcement of
Brewer's result, in which ''distributed databases'' appear as an
example of unavailable systems~\cite{brewer-slides}. While, in the
same document, Brewer is careful to note that consistency and
availability are a ``spectrum,'' all of ACID is still often labeled
with a binary categorization of ``unavailable''~\cite{hn,
  foundation-article}.

@@

It is unknown \textit{which} of these ACID (particularly isolation)
guarantees can be provided with high availability. The existing
algorithms for providing weak isolation are typically designed for a
single-node context and are, to the best of our knowledge, unavailable
due to a reliance on concurrency control mechanisms like locking that
are not resilient to partial failure
(Section~\ref{sec:evaluation}). Morever, we are not aware of any prior
literature that provides guidance as to the relationship between weak
isolation and high availability: prior work has examined the
relationship between serializability and high
availability~\cite{davidson-survey} and weak isolation on a
single server~\cite{adya, ansicritique, gray-isolation} but never weak isolation and
high availability taken together.

### 2013-03-31 writeup/relatedwork.tex
checkpoint

>>>>

Wisemann et al. propose a three parameter classification for one-copy
serializable database systems, performing a more extensive
classification than our overview in Setion XX, including eager
database replication~\cite{kemme-classification}. Additional, several
surveys discuss distributed mechanisms for achieving one-copy
serializability~\cite{wisemann-survey}, linearizability, consistent
snapshots, and recency bounds~\cite{ceri-mechanism, chen-mechanism}.

% Ceri discuss distributed database update mechanisms with respect to
% linearizability, consistent snapshots (i.e., cut isolation), and
% recency bounds~\cite{ceri-mechanism}.  Chen and Pu also classify
% distributed replica maintenance from perspectives of
% linearizability,% recency guarantees, and replica
% divergence~\cite{chen-mechanism}.

, with increasing interest in returning to transactional
designs~\cite{spanner, walter, foundation-article, krikellas-bargain,
  eiger}
